---
title: "CSV Summary Analysis"
output:
  html_document:
    theme: bootstrap
    toc: true
    toc_depth: 3
---
This R Markdown document reads, combines, and summarizes multiple
`_summary.csv` files.
***
## 0. Packages

-[DESCRIBE CODE AND VARIABLES]-

 -tidyverse- Provides `dplyr`, `readr`, `purrr`, `stringr`, and `ggplot2`, which are used throughout this analysis for data import, manipulation, and summarization.
```{r setup, message=FALSE}
library(tidyverse)

# Ensure rmarkdown can find Pandoc.
# On Windows, Quarto bundles pandoc in: <quarto>/bin/tools/pandoc.exe.
if (!nzchar(Sys.which("pandoc"))) {
  quarto_exe <- Sys.which("quarto")
  if (nzchar(quarto_exe)) {
    quarto_tools <- normalizePath(
      file.path(dirname(quarto_exe), "tools"),
      winslash = "/",
      mustWork = FALSE
    )
    if (dir.exists(quarto_tools)) {
      Sys.setenv(RSTUDIO_PANDOC = quarto_tools)
    }
  }
}
```
***

## 1. Read and combine ALL CSVs

-[DESCRIBE CODE AND VARIABLES]-

 -folder- Path to the directory containing all summary CSV files.
 -files- Vector of file paths matching the `_summary.csv` naming pattern.
 -raw_list- List of data frames, one per CSV file.
```{r import-csv}
folder <- "C:/Users/dunnmk/Downloads/C_summary"
files  <- list.files(folder, pattern = "_summary\\.csv$", full.names = TRUE)
raw_list <- lapply(files, read_csv)
dat <- bind_rows(raw_list)
str(dat)
dat |> head(10) |> knitr::kable(caption = "Combined CSV preview")

```
***

## 2. Aggregate counts and percentages


Purpose- Calculate percentages from aggregated counts

```{r aggregate-counts}
dat_raw <- dat
summarize_counts_pct <- function(dat){
  dat %>%
    summarise(
      Total_Counts   = sum(count, na.rm = TRUE),
      Cis_Counts     = sum(count[cis_trans == "CIS"], na.rm = TRUE),
      Trans_Counts   = sum(count[cis_trans == "TRANS"], na.rm = TRUE),
      A_to_B_Counts  = sum(count[combo == "A_to_B"], na.rm = TRUE),
      C_to_D_Counts  = sum(count[combo == "C_to_D"], na.rm = TRUE),
      A_to_D_Counts  = sum(count[combo == "A_to_D"], na.rm = TRUE),
      C_to_B_Counts  = sum(count[combo == "C_to_B"], na.rm = TRUE),
      Percent_Cis    = if_else(Total_Counts > 0, 100 * Cis_Counts / Total_Counts, NA_real_),
      Percent_Trans  = if_else(Total_Counts > 0, 100 * Trans_Counts / Total_Counts, NA_real_),
      Percent_A_to_B = if_else(Total_Counts > 0, 100 * A_to_B_Counts / Total_Counts, NA_real_),
      Percent_C_to_D = if_else(Total_Counts > 0, 100 * C_to_D_Counts / Total_Counts, NA_real_),
      Percent_A_to_D = if_else(Total_Counts > 0, 100 * A_to_D_Counts / Total_Counts, NA_real_),
      Percent_C_to_B = if_else(Total_Counts > 0, 100 * C_to_B_Counts / Total_Counts, NA_real_),
      .groups = "drop"
    )
}
dat <- dat %>%
  group_by(batch, time_point, DSB) %>%
  summarize_counts_pct()
str(dat)
```
Print data after aggregation and new columns

xxxxx
***

# 2. cis/trans normalization — compute totals and percentages per allele

```{r cis-trans-norm}
my_summarize_cistrans <- function(dat){
  by_allele <- dat %>%
    group_by(batch, time_point, DSB, allele) %>%
    summarise(
      Cis_Location_Counts   = sum(count[cis_trans == "CIS"], na.rm = TRUE),
      Trans_Location_Counts = sum(count[cis_trans == "TRANS"], na.rm = TRUE),
      .groups = "drop"
    )

  totals <- by_allele %>%
    group_by(batch, time_point, DSB) %>%
    summarise(
      Total_Cis_Location_Counts   = sum(Cis_Location_Counts, na.rm = TRUE),
      Total_Trans_Location_Counts = sum(Trans_Location_Counts, na.rm = TRUE),
      .groups = "drop"
    )

  by_allele %>%
    left_join(totals, by = c("batch", "time_point", "DSB")) %>%
    mutate(
      Percent_Location_in_Cis   = if_else(Total_Cis_Location_Counts > 0,
                                         100 * Cis_Location_Counts / Total_Cis_Location_Counts,
                                         NA_real_),
      Percent_Location_in_Trans = if_else(Total_Trans_Location_Counts > 0,
                                         100 * Trans_Location_Counts / Total_Trans_Location_Counts,
                                         NA_real_)
    )
}

dat_norm <- my_summarize_cistrans(dat_raw)
dat_norm |> head(10) |> knitr::kable(caption = "Data after cis/trans normalization")
```
***
# 3. Allele frequency (CIS + TRANS)

```{r allele-freq}
my_summarize_allelefreq <- function(dat){
  by_allele <- dat %>%
    group_by(batch, time_point, DSB, allele) %>%
    summarise(
      Cis_Counts    = sum(count[cis_trans == "CIS"], na.rm = TRUE),
      Trans_Counts  = sum(count[cis_trans == "TRANS"], na.rm = TRUE),
      Allele_Total  = Cis_Counts + Trans_Counts,
      .groups = "drop"
    )

  totals <- by_allele %>%
    group_by(batch, time_point, DSB) %>%
    summarise(
      Total_Cis   = sum(Cis_Counts, na.rm = TRUE),
      Total_Trans = sum(Trans_Counts, na.rm = TRUE),
      Total_All   = Total_Cis + Total_Trans,
      .groups = "drop"
    )

  by_allele %>%
    left_join(totals, by = c("batch", "time_point", "DSB")) %>%
    mutate(
      Allele_Frequency = if_else(Total_All > 0, Allele_Total / Total_All, NA_real_)
    )
}

dat_allele_freq <- my_summarize_allelefreq(dat_raw)
str(dat_allele_freq)

```
***
# 4. Fold change calculations (120 / 0)

```{r fold-change}
library(tidyr)

dat_fc_cis <- dat_norm %>%
  filter(time_point %in% c(0, 120)) %>%
  select(batch, time_point, DSB, allele, Percent_Location_in_Cis) %>%
  pivot_wider(names_from = time_point, values_from = Percent_Location_in_Cis, values_fill = 0) %>%
  mutate(
    FoldChange_Cis_120_vs_0 = `120` / `0`,
    Log2FC_Cis_120_vs_0 = log2(`120` / `0`)
  )

dat_fc_trans <- dat_norm %>%
  filter(time_point %in% c(0, 120)) %>%
  select(batch, time_point, DSB, allele, Percent_Location_in_Trans) %>%
  pivot_wider(names_from = time_point, values_from = Percent_Location_in_Trans, values_fill = 0) %>%
  mutate(
    FoldChange_Trans_120_vs_0 = `120` / `0`,
    Log2FC_Trans_120_vs_0 = log2(`120` / `0`)
  )

```
***
# 5. Correlation: log2 Fold Change vs Allele Frequency

```{r cor-allelefreq}
library(tidyr)

dat_wide <- dat_norm %>%
  filter(time_point %in% c(0, 120)) %>%
  select(batch, DSB, allele, time_point, Percent_Location_in_Cis, Percent_Location_in_Trans) %>%
  pivot_wider(names_from = time_point, values_from = c(Percent_Location_in_Cis, Percent_Location_in_Trans), values_fill = 0) %>%
  mutate(
    log2FC_CIS   = log2((Percent_Location_in_Cis_120 + 1e-6) / (Percent_Location_in_Cis_0 + 1e-6)),
    log2FC_TRANS = log2((Percent_Location_in_Trans_120 + 1e-6) / (Percent_Location_in_Trans_0 + 1e-6))
  )

dat_fc_af <- dat_wide %>%
  inner_join(
    dat_allele_freq %>% filter(time_point == 120) %>% select(batch, DSB, allele, Allele_Frequency),
    by = c("batch", "DSB", "allele")
  ) %>%
  filter(!is.na(Allele_Frequency) & Allele_Frequency > 0)

cor_summary <- dat_fc_af %>%
  group_by(batch, DSB) %>%
  summarise(
    cor_CIS_AF   = ifelse(n() >= 2, cor(log2FC_CIS, Allele_Frequency), NA),
    cor_TRANS_AF = ifelse(n() >= 2, cor(log2FC_TRANS, Allele_Frequency), NA),
    n_obs = n()
  )

knitr::kable(cor_summary, caption = "Correlation: log2FC vs Allele Frequency")

```
***
# 6a. CIS percentage bar plot 

```{r CIS-bar-plot}
dat_plot_cis <- dat_raw %>%
  filter(time_point %in% c(0, 120)) %>%
  group_by(batch, time_point, DSB, allele) %>%
  summarize_counts_pct()

ggplot(dat_plot_cis, aes(x = allele, y = Percent_Cis)) +
  geom_col(width = 0.9, fill = "darkgreen") +
  facet_grid(time_point ~ batch, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 8)) +
  labs(title = "Percent CIS by allele (T0 + T120)", x = "Allele", y = "Percent CIS")

```
# 6b. trans percentage bar plot 

```{r trans-bar-plot}
# 6b. TRANS percentage bar plots (A_to_D, C_to_B)
# Built directly from filtered subsets of the original table, using the same summarise helper.

plot_trans_percent <- function(dat, combo_name) {
  df_plot <- dat %>%
    filter(combo == combo_name, time_point %in% c(0, 120)) %>%
    group_by(batch, time_point, DSB, allele) %>%
    summarize_counts_pct()

  if (nrow(df_plot) > 0) {
    ggplot(df_plot, aes(x = allele, y = Percent_Trans)) +
      geom_col(fill = "steelblue", width = 0.9) +
      facet_grid(time_point ~ batch, scales = "free_x") +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 8)) +
      labs(
        title = paste("Percent TRANS by allele:", combo_name, "(T0 + T120)"),
        x = "Allele", y = "Percent TRANS"
      )
  }
}

# Generate both TRANS plots
plot_trans_percent(dat_raw, "A_to_D")
plot_trans_percent(dat_raw, "C_to_B")


```
***
# 7. Fold Change bar plot 

```{r foldchange-bar-plot}
plot_foldchange_cis <- function(dat_fc_cis) {
  ggplot(dat_fc_cis, aes(x = allele, y = FoldChange_Cis_120_vs_0)) +
    geom_col(width = 0.9, fill = "orange") +
    facet_grid(DSB ~ batch, scales = "free_x") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 8)) +
    labs(title = "Fold change (120/0) in CIS by allele", x = "Allele", y = "Fold change (120/0)")
}

plot_foldchange_trans <- function(dat_fc_trans) {
  ggplot(dat_fc_trans, aes(x = allele, y = FoldChange_Trans_120_vs_0)) +
    geom_col(width = 0.9, fill = "steelblue") +
    facet_wrap(~ batch, scales = "free_x") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 8)) +
    labs(title = "Fold change (120/0) in TRANS by allele", x = "Allele", y = "Fold change (120/0)")
}

plot_allele_frequency <- function(dat_allele_freq) {
  ggplot(dat_allele_freq, aes(x = allele, y = Allele_Frequency)) +
    geom_col(width = 0.9, fill = "purple") +
    facet_grid(time_point ~ batch, scales = "free_x") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 0.5)) +
    labs(title = "Allele Frequency (CIS + TRANS)", x = "Allele", y = "Allele Frequency")
}

# Generate fold change plots
plot_foldchange_cis(dat_fc_cis)
plot_foldchange_trans(dat_fc_trans)
plot_allele_frequency(dat_allele_freq)

# CIS fold change plot
p1 <- ggplot(dat_fc_cis, aes(x = allele, y = FoldChange_Cis_120_vs_0, fill = DSB)) +
  geom_col(width = 0.9, position = "dodge") +
  facet_wrap(~ batch, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 8)) +
  labs(title = "Fold change (120/0) CIS by allele", x = "Allele", y = "Fold change")

print(p1)

# TRANS fold change plot  
p2 <- ggplot(dat_fc_trans, aes(x = allele, y = FoldChange_Trans_120_vs_0, fill = DSB)) +
  geom_col(width = 0.9, position = "dodge") +
  facet_wrap(~ batch, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 8)) +
  labs(title = "Fold change (120/0) TRANS by allele", x = "Allele", y = "Fold change")

print(p2)


```
***
# 8. Correlation scatter plots (log2FC vs Allele Frequency)
```{r cor-allele-log}
library(ggrepel)  # Add this line

plot_correlation_cis <- function(dat_fc_af, batch_name) {
  df_batch <- dat_fc_af %>% filter(batch == batch_name, DSB == "DSB1")
  if (nrow(df_batch) >= 2) {
    ggplot(df_batch, aes(x = Allele_Frequency, y = log2FC_CIS, label = allele)) +
      geom_point(size = 3, alpha = 0.8, color = "darkgreen") +
      geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
      geom_smooth(method = "lm", se = FALSE, color = "red") +
      geom_text_repel(size = 3, max.overlaps = 20) +
      theme_bw() +
      labs(
        title = paste0("Log2FC CIS vs Allele Freq | Batch: ", batch_name),
        x = "Allele Frequency", y = "log2FC CIS"
      )
  }
}

plot_correlation_trans <- function(dat_fc_af, batch_name) {
  df_batch <- dat_fc_af %>% filter(batch == batch_name)
  if (nrow(df_batch) >= 2) {
    ggplot(df_batch, aes(x = Allele_Frequency, y = log2FC_TRANS, color = DSB, label = allele)) +
      geom_point(size = 3, alpha = 0.8) +
      geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
      geom_smooth(method = "lm", se = FALSE) +
      geom_text_repel(size = 3, max.overlaps = 20) +
      theme_bw() +
      labs(
        title = paste0("Log2FC TRANS vs Allele Freq | Batch: ", batch_name),
        x = "Allele Frequency", y = "log2FC TRANS"
      )
  }
}

# Generate plots
unique_batches <- unique(dat_fc_af$batch)
for (b in unique_batches) {
  plot_correlation_cis(dat_fc_af, b)
  plot_correlation_trans(dat_fc_af, b)
}
```
